### 2018-12-07 paper/sex_selection_spacing-ver2.tex
Add Appendix discussion of recall error

>>>>

\section{Recall Error and the Sex Ratio}

The reliability of the results depends on the correctness of the birth histories
provided by the respondents.
A significant concern here is underreporting of child mortality, especially a systematic
recall error where respondents' likelihood of reporting a deceased child depends on the 
sex of that child. 
This section assesses the degree of recall error across the surveys and discusses methods
to address it.

NFHS enumerators probe for any missed births, although the method depends on the survey.
NFHS-1 probe for each calendar birth interval that is 4 or more years.
NFHS-2 asked for stillbirths, spontaneous and induced abortions and also probed 
for each calendar birth interval 4 or more years.
NFHS-3 and NFHS-4 did not directly use birth intervals, but asked whether were there any 
other live births between (name of previous birth) and (name), including any children who 
died after birth, and asked for births before the birth listed as first birth and
after the last birth listed as the last birth.

Probing catches many initially missed births, but systematic recall error based on son
preference may still be a problem.
First, son preference leads to significantly higher mortality for girls than boys.
Secondly, son preference makes it more likely that parents will remember deceased boys 
than deceased girls.
Finally, in the absence of sex-selective abortions, parents with a preference for sons may
have the next birth sooner if the last child was a girl than if it was a boy.
If this girl subsequently dies, she is more likely to be missed if probing for missed 
births is only done for long intervals as in NFHS-1 and NFHS-2.

I use two approaches to examine the degree of recall error.
The first approach is to test whether the observed sex ratio is significantly different
from the natural sex ratio.
Prenatal sex determination techniques did not become widely available until the mid-1980s, 
so any significant deviation from the natural sex ratio before that time is likely the 
result of recall error.
The second approach is to compare births that took place during the same period but
where captured in different surveys.
Recall error is likely to increase with time, so births and deaths that took place earlier 
are more likely to be subject to recall error than more recent events.

Table \ref{tab:recallBirthBO1} shows the sex ratios of children recorded as first-born by 
year of birth, together with tests for whether the observed sex ratio is significantly 
higher than the natural sex ratio and whether more recent surveys have a higher sex ratio 
for the cohort than earlier surveys for the same period births.
Births are combined into five-year cohorts to achieve sufficient power.

\input{../tables/recallBirthBO1.tex}

% \input{../tables/recallBirthBO2.tex}

The ``first-born'' sex ratios illustrate the systematic recall error problem well.
In all four surveys around 55 percent of children reported as first-born are boys
for the first cohort of births observed.
Given that these cohorts cover from 1960-1964 to 1980-1984, which is before sex selection 
techniques became available in India, the most likely explanation for the skewed sex ratio 
is that some children listed as first-borns were not, in fact, the first children born in 
their families.
Instead, for a substantial proportion of families, their first-born was a girl who died 
and went unreported when enumerators asked about birth history.

As expected, the difference between the observed sex ratio and the natural sex ratio is 
less pronounced the closer to the survey date the cohort is.
The observed sex ratio for children born just before the NFHS-1 survey and listed as 
first-born is 0.517, which is not statistically significantly different from the
natural sex ratio.
The same general pattern holds for the other three surveys, with cohorts further away
from the survey date more likely to have a sex ratio skewed male.

% Second births show a pattern very similar to that for first births.
% An interesting difference is that there is evidence of a U-shaped relationship between
% time and sex ratio for second births.
% Cohorts furthest away from the survey year show the highest sex ratios, but declines to the
% natural sex ratio in the mid-1980s, and the sex ratio is then significantly higher again 
% for more recent births.
% This is in line with the results in the main paper that show that sex selective abortions 
% take place on lower parity births as desired fertility declined.

Finally, across surveys, the same cohort tends to show a higher sex ratio the more recent 
the survey (births in the cohort took place earlier relatively to the survey date).
Despite this, few cohorts show significantly different sex ratios across surveys, most 
likely because of a lack of power.
The exception is that comparisons involving NFHS-4 are mostly statistically significant
since the number of surveyed households in NFHS-4 were much larger than in prior surveys.

The problem with the above approach is that the year of birth is affected by recall error; 
for example, a second born child listed as first-born is born later than the real 
first born child.
Recorded year of marriage should, however, be affected neither by parental recall error 
nor the use of sex-selective abortions.
Tables \ref{tab:recallMarriageBO1} and \ref{tab:recallMarriageBO2}, therefore, shows sex 
ratios of children recorded as first-born and second-born by year of parents' marriage, 
together with tests for whether the observed sex ratio is significantly higher than the 
natural sex ratio and whether more recent surveys show a higher sex ratio for the cohort 
than earlier surveys.

The basic recall error pattern remains with women married longer ago more
likely to report that their first-born is a boy.
If anything the pattern is more consistent when using year of marriage than 
year of birth. 
Before the introduction of sex selection there are only two instances  across the two 
parities where the observed sex ratio is lower for women married longer ago than for women 
married closer to the survey date.%
\footnote{
For first-borns it is between 1970--1974 and 1975-1979 for NFHS-3 and for second-borns
it is between 1975--1979 and 1980--1984 for NFHS-1.
}
Similarly, comparing women married in the same five-year period across surveys shows
that women married longer ago are more likely to report having a son.

\input{../tables/recallMarriageBO1.tex}

\input{../tables/recallMarriageBO2.tex}

The relationship between length of marriage and recall error can also seen in 
Figure \ref{fig:sexRatioMarriage}, which shows the observed sex ratio for children reported
as first born as a function of duration of marriage at the time of the survey combining 
all four surveys.
The solid line is the sex ratio of children reported 
as first born by the number of years between the survey and marriage, 
the dashed lines indicate the 95 percent confidence interval 
and the horizontal line the natural sex ratio (approximately 0.512).
To ensure sufficient cell sizes the years are grouped in twos.
In line with the results from Tables \ref{tab:recallMarriageBO1} and 
\ref{tab:recallMarriageBO2}, the observed ratio of boys 
is increasingly above the expected value the longer ago the parents were married.%
\footnote{
The graph for second births shows a similar pattern with the
likelihood of the second child being a boy going up with increasing marriage duration.
The graphs for the second births and the individual survey rounds are available upon request.
}

\begin{figure}[htp]%
\centering
\includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
\caption{Sex ratio of ``first'' births by duration of marriage}
\label{fig:sexRatioMarriage}}
\end{figure}

The increasingly unequal sex ratio with increasing marriage duration suggests that
a solution to the recall error problem is to drop observations for 
women who were married ``too far'' from the survey year.
The main problem is establishing what the best cut-off point should be.
As Tables \ref{tab:recallMarriageBO1} and \ref{tab:recallMarriageBO2} shows there are 
differences in recall error across the three surveys and between parities.
One possibility is simply to drop all women married in cohorts where the observed
sex ratio is substantially higher than the natural sex ratio, but even this is not
straightforward.
For ``first'' born this would imply dropping all women married before 1980, but
for ``second'' born it would be imply dropping all married before 1970 for NFHS-1, before 1975 
for NFHS-2, and before 1980 NFHS-3.
Because of the differences between ``first'' and ``second'' born in the different surveys
I vary the cut-off point by survey round.

NFHS-3 is the most straightforward in that the implied cut-off point is the same for
both first and second born.
Based on Tables \ref{tab:recallMarriageBO1} and \ref{tab:recallMarriageBO2}, women in 
NFHS-3 who were married 26 years or
more before the survey date are dropped.
For NFHS-1 and NFHS-2 the trade-off is between ensuring enough information and not bias 
the results unduly.
Given that sex selective abortions are more likely in higher parity birth I base the
cut-off point on the second births instead of the first births.
Hence, for NFHS-1 women who were married 22 years or more before the survey date 
were dropped and for NFHS-2 women who were married 23 year or more before the survey date 
were dropped.

% \begin{figure}
% \centering
% \subfloat[NFHS-1]{\includegraphics[width=.49\textwidth]{des_b1_m_age_r1}}
% \subfloat[NFHS-2]{\includegraphics[width=.49\textwidth]{des_b1_m_age_r2}} \\
% \subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{des_b1_m_age_r3}} 
% \caption{Ratio of Boys for ``First'' Births by Survey Round}
% \label{fig:sex_ratio_recall_rounds}
% \end{figure}
